Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-19 co-regular-EPQ.tex
modify

Register automata is a kind of finite automata with registers used to deal with infinite data domain, where register can be assigned values and do simple operations. In this paper we use a restricted kind of register automata, which contains only one register, whose value is obtained once by guessing and never changed. Our register machine contains a register $r$. Its transition labels are $r=*,\textit{call}(\textit{put},d,g), \textit{ret}(\textit{put},d,g),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards. The first transition from initial state is a $r=*$ transition, and is used to assign an arbitrary value to $r$. A guard is a predicate chosen from $\{=r,\leq r,\top \}$, which represents equals the value of $\textit{reg}$, less than the value of $\textit{reg}$, and always holds, respectively. Given an execution $e = \alpha_1 \cdots \alpha_k$ of priority queue, register automata $\mathcal{A}$ accepts $e$, if there exists value $d_r\in \mathbb{D}$ and transitions $q_0 \xrightarrow{\beta_1} q_1 \ldots \xrightarrow{\beta_k} q_k$ of $\mathcal{A}$ from a initial state to a final state, such that for each $i$

@@

%\vspace{-6pt}
\begin{definition}\label{def:co-regular of rules of extended priority queues}
Given $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$. $\Gamma$ is co-regular, if there is a finite set $\textit{Auts}_{\Gamma}$ of register automata, such that for each data-independence implementation $\mathcal{I}$, we have that:

@@

We use two point to simply our proof co-regular. The first point is to ues the results in \cite{Bouajjani:2015} to ensure FIFO (first in first out) property of each single-priority projections. A execution where only values of one priority is putted is called a single-priority execution. It is obvious that each single-priority projection of sequences in $\seqPQ$ satisfy FIFO property. Let $\textit{toQueue}(e)$ be an execution generated from $e$ by transforming $\textit{put}$ and $\textit{rm}$ into $\textit{enq}$ and $\textit{deq}$, respectively, and then discarding priorities. In Appendix \ref{sec:appendix proof and definition in section definition of co-regular}, according to the result of queue in \cite{Bouajjani:2015}, we construct a set $\textit{Auts}_{\textit{sinPri}}$ of witness automata, and shows that they are enough to ensure that for each data-differentiated executions, each of its single-priority projection without $\textit{rm}(\textit{empty})$ to have ``FIFO'' property, as shown by the following lemma.

@@

The second point is to ignore the incomparable priorities. Given a data-differentiated execution $e$ and one of its maximal priority $p$, $e\vert_{\preceq p}$ is an execution obtained from $e$ by erasing operations of items with priority incomparable with $p$. The following lemma states that when checking co-regular of $\seqPQ$, it is safe to ignore all incomparable priorities.

@@

To prove the $\textit{only if}$ direction, given $e \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \Gamma\mathsf{\text{-}Seq}(s,\alpha)$. Since there is no restriction to $p$-incomparable operations in $u$, $v$ and $w$, we can safely erase all $p$-incomparable values.

To prove the $\textit{if}$ direction, we use a lemma in Appendix, which said that if the projection an execution into a set $O_1$ of operations has linearization $l_1$ and the projection of the execution into $O_2$, the remanning operation, has linearization $l_2$, then the execution has a linearization whose projection into $O_1$ is $l_1$ and projection into $O_2$ is $l_2$.

Given $e\vert_{\preceq p} \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \Gamma\mathsf{\text{-}Seq}(s,\alpha)$. By above lemma we know that $e$ has a linearization $l'$, such that the projection of $l'$ into $p$-incomparable values. It is not hard to see that the priorities comparable with $p$ are either smaller or incomparable with the priorities incomparable with $p$. Thus, $l' \in \seqPQ$. \qed
\end {proof}

Lemma \ref{lemma:pri execution is enough} makes our co-regular proof more clear. One consequence of it is that for $e=\textit{p-Exec}(e)$, at most one $\Gamma\mathsf{\text{-}Condition}(e')$ holds.

@@

To facilitate the proof of co-Regular of $\mathsf{MatchedMaxPriority}$, we separate $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ into the following two sub-predicates, where the first one represents that $x$ is the only value with priority $p$ in $e$ , and the second one represents that there are more than one value with priority $p$ in $e$. Here $p\not\prec'\textit{priorities}(l)$ means that $p$ is either larger, equal, or incomparable with priorities in $l$, and $p$ must equals with some priority in $l$.

{\small
\begin{align*}
\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v\cdot \textit{rm}(x)\cdot w$, $p\not\preceq \textit{priorities}(u\cdot v\cdot w)$,} \\
&\hspace{4mm}\mbox{$p\not\preceq \textit{unmatched-priorities}(u\cdot v\cdot w)$, $\textit{matched}_\prec(u\cdot v,p)$,} \\
&\hspace{4mm}\mbox{and $p\not\in \textit{priorities}(v\cdot w)$} \\
\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v\cdot \textit{rm}(x)\cdot w$, $p\not\prec' \textit{priorities}(u\cdot v\cdot w)$,} \\
&\hspace{4mm}\mbox{$p\not\preceq \textit{unmatched-priorities}(u\cdot v\cdot w)$, $\textit{matched}_\prec(u\cdot v,p)$,} \\
&\hspace{4mm}\mbox{and $p\not\in \textit{priorities}(v\cdot w)$} 
\end{align*}
}

The corresponding conditions are defined as follows: $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$

{\small
\begin{align*}
\mathsf{HasEmptyRemoves}(e)=\mathsf{true} & \mbox{ iff  $e$ contains a $\textit{rm}(\textit{empty})$ operation} \\
\mathsf{HasUnmatchedMaxPriority}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{unmatched-priorities}(e)$ for a maximal priority} \\
&\hspace{4mm}\mbox{$p\in priorities(e)$}
\end{align*}}

@@

@@

$\Gamma\mathsf{\text{-}Condition}(e')$ holds but $\Gamma\mathsf{\text{-}Conc}(e',\alpha)$ does not hold

Given a data-differentiated $\_$-execution $e$ such that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$ holds, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensuring that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ for some $x$. This is because that it is possible that $e \not\sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$ because of interaction between actions of multiple priorities. One such example is shown in \figurename~\ref{fig:introduce gap for EPQ1Lar}. We call the time interval from $\textit{ret}(\textit{put},x)$ to $\textit{cal}(rm,x)$, or from $\textit{ret}(\textit{put},x)$ when $\textit{cal}(rm,x)$ does not exist, the interval of item $x$. In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each item by dashed line. Here we assume that $p_1 <_{\mathbb{P}} p_4$, $p_2 <_{\mathbb{P}} p_4$ and $p_3 <_{\mathbb{P}} p_4$. The reason of $e \not\sqsubseteq \textit{MS}(\textit{PQ}_1^{>})$ is that, each time point from $\textit{cal}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is in interval of some item with smaller priority.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that does not linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we introduce the notion of left-right constraint, which is a graph, and the existence of cycle though item with maximal priority in it formally modelled the case in \figurename~\ref{fig:introduce gap for EPQ1Lar}. When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through item $x$ with maximal priority in left-right constraint, we say that $x$ is covered by $d_1,\ldots,d_m$ For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $b$ is covered by $f$, $e$ and $d$. We then prove that getting rid of cycle though item with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$, as stated by the following lemma:

\begin{restatable}{lemma}{LinEqualsConstraintforEPQOneLar}
\label{lemma:Lin Equals Constraint for EPQ1Lar}
Given a data-differentiated $\textit{pri}_x$-execution $e$ with $\textit{last}(e) = \textit{EPQ}_1^{>}$. Let $\textit{put}(x,\textit{pri}_x)$ and $\textit{rm}(x)$ be method events of $e$ with maximal priority. Let $G$ be the graph representing the left-right constraint of $\textit{put}(x)$ and $\textit{rm}(x)$. $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$, if and only if $G$ has no cycle going through $x$.
\end{restatable}

Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other item into $c$ by data-independence. Such execution can be recognized by witness automata, since between the first $\textit{ret}(\textit{put},a,\_)$ and the last $\textit{cal}(\textit{rm},a)$ (if exists), $\textit{ret}(\textit{put},a,\_)$ and $\textit{cal}(\textit{rm},a)$ occurs in pair and there is no need for count. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we construct a set $\textit{Auts}_{\textit{1-lar}}$ of witness automata, and use $\textit{Auts}_{\textit{1-lar}}$ to prove that $\textit{EPQ}_1^{>}$ is co-regular, as stated by the following lemma.

\begin{restatable}{lemma}{EPQOneLarisCoRegular}
\label{lemma:EPQ1Lar is co-regular}
$\textit{EPQ}_1^{>}$ is co-regular.
\end{restatable}

>>>>

Register automata is a kind of finite automata with registers used to deal with infinite data domain, where register can be assigned values and do simple operations. In this paper we use a restricted kind of register automata, which contains only one register, whose value is obtained once by guessing and never changed. Our register machine contains a register $r$. Its transition labels are $r=*,\textit{call}(\textit{put},d,g), \textit{ret}(\textit{put},d,g),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards. The first transition from initial state is a $r=*$ transition, and is used to assign an arbitrary value to $r$. A guard is a predicate chosen from $\{=r,\leq r,\top \}$, which represents equals the value of $r$, less than the value of $r$, and always holds, respectively. Given an execution $e = \alpha_1 \cdots \alpha_k$ of priority queue, register automata $\mathcal{A}$ accepts $e$, if there exists value $d_r\in \mathbb{D}$ and transitions $q_0 \xrightarrow{\beta_1} q_1 \ldots \xrightarrow{\beta_k} q_k$ of $\mathcal{A}$ from a initial state to a final state, such that for each $i$

@@

%\vspace{-6pt}
\begin{definition}\label{def:co-regular of rules of extended priority queues}
Given $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$. $\Gamma$ is co-regular, if there is a finite set $\textit{Auts}_{\Gamma}$ of register automata, such that for each data-independent implementation $\mathcal{I}$, we have that:

@@

We use two point to simply our proof of co-regular. The first point is to use the results in \cite{Bouajjani:2015} to ensure FIFO (first in first out) property of each single-priority projections. A execution where only values of one priority is putted is called a single-priority execution. It is obvious that each single-priority projection of sequences in $\seqPQ$ satisfy FIFO property. Let $\textit{toQueue}(e)$ be an execution generated from $e$ by transforming $\textit{put}$ and $\textit{rm}$ into $\textit{enq}$ and $\textit{deq}$, respectively, and then discarding priorities. In Appendix \ref{sec:appendix proof and definition in section definition of co-regular}, according to the result of queue in \cite{Bouajjani:2015}, we construct a set $\textit{Auts}_{\textit{sinPri}}$ of witness automata, and shows that they are enough to ensure that for each data-differentiated executions, each of its single-priority projection without $\textit{rm}(\textit{empty})$ to have ``FIFO'' property, as shown by the following lemma.

@@

The second point is to ignore the incomparable priorities. Given a data-differentiated execution $e$ and one of its maximal priority $p$, $e\vert_{\preceq p}$ is an execution obtained from $e$ by erasing operations of values with priority incomparable with $p$. The following lemma states that when checking co-regular of $\seqPQ$, it is safe to ignore all incomparable priorities.

@@

To prove the $\textit{only if}$ direction, given $e \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. Since there is no restriction to $p$-incomparable operations in $u$, $v$ and $w$, we can safely erase all $p$-incomparable values. Similar case holds for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

To prove the $\textit{if}$ direction, we use a lemma in Appendix, which said that if the projection of an execution into a set $O_1$ of operations has linearization $l_1$ and the projection of the execution into $O_2$, the remanning operations, has linearization $l_2$, then the execution has a linearization whose projection into $O_1$ is $l_1$ and projection into $O_2$ is $l_2$.

Given $e\vert_{\preceq p} \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. By above lemma we know that $e$ has a linearization $l' = u' \cdot \textit{put}(x,p) \cdot v' \cdot \textit{rm}(x) \cdot w'$, such that the projection of $l'$ into $p$-incomparable values is l. We can see that $u$, $v$ and $w$ are sub-sequence of $u'$, $v'$ an $w'$, respectively. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$ does not have restriction to $p$-incomparable values, we can see that $l' \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. \qed
\end {proof}

Lemma \ref{lemma:pri execution is enough} makes our co-regular proof more clear. One consequence of it is that for given $e\vert_{\preceq p}$, $\Gamma\mathsf{\text{-}Condition}(e')$ holds only for one $\Gamma$.

@@

To facilitate the proof of co-Regular of $\mathsf{MatchedMaxPriority}$, we separate $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ into the following two sub-predicates, where the first one represents that $x$ is the only value with priority $p$ in $e$ , and the second one represents that there are more than one value with priority $p$ in $e$. Here $\textit{occurOnce}(p,l)$ holds if there is only one value with priority $p$ in $l$.

{\small
\begin{align*}
\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
&\hspace{4mm}\mbox{$\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
&\hspace{4mm}\mbox{$\neg\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
\end{align*}
}

@@

We can also define two conditions, where $\textit{unmatched-priorities}(l)$ is the set of priorities occurring in $\textit{put}$ operations of $e$ for which there is $\textit{rm}$ operation removing the same value:

{\small
\begin{align*}
\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
&\hspace{4mm}\mbox{$p\in priorities(e)$, $\textit{occurOnce}(p,e)$} \\
\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
&\hspace{4mm}\mbox{$p\in priorities(e)$, $\neg\textit{occurOnce}(p,e)$} 
\end{align*}}

@@

Given a data-differentiated $\_$-execution $e$ such that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$ holds, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensuring that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ for some $x$. This is because that it is possible that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ does not holds because of interaction between actions of multiple priorities. One such example is shown in \figurename~\ref{fig:introduce gap for EPQ1Lar}. We call the time interval from $\textit{ret}(\textit{put},x)$ to $\textit{cal}(rm,x)$, or from $\textit{ret}(\textit{put},x)$ when $\textit{cal}(rm,x)$ does not exist, the interval of value $x$. In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each value by dashed line. Here we assume that $p_1 \prec p_4$, $p_2 \prec p_4$ and $p_3 \prec p_4$. We can not find a sequence $s$ where $e \sqsubseteq s$ and $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(s,b)$ holds, since each time point from $\textit{cal}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is in interval of some item with smaller priority, and $\textit{rm}(b)$ can not happen in interval of item with smaller priority.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that does not linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we introduce the notion of left-right constraint, which is a graph, and the existence of cycle though item with maximal priority in it formally modelled the case in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $b$ is covered by $f$, $e$ and $d$. We then prove that getting rid of cycle though item with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$, or we can say, ensure $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ for some $x$.

When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through item $x$ with maximal priority in left-right constraint, we say that $x$ is covered by $d_1,\ldots,d_m$. Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other item into $c$ by data-independence. Such execution can be recognized by witness automata, since between the first $\textit{ret}(\textit{put},a,\_)$ and the last $\textit{call}(\textit{rm},a)$ (if exists), $\textit{ret}(\textit{put},a,\_)$ and $\textit{call}(\textit{rm},a)$ occurs in pair and there is no need for count. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we construct a set $\textit{Auts}_{\textit{1-lar}}$ of witness automata, and use $\textit{Auts}_{\textit{1-lar}}$ to prove that $\mathsf{MatchedMaxPriority}^{>}$ is co-regular, as stated by the following lemma.

\begin{restatable}{lemma}{EPQOneLarisCoRegular}
\label{lemma:EPQ1Lar is co-regular}
$\mathsf{MatchedMaxPriority}^{>}$ is co-regular.
\end{restatable}
